Short and Mechanized Logical Relation for Dependent Type Theory

### 2024-02-21 paper/paper.tex
Add explanation for valuation

The logical relation we define in Figure~\ref{fig:logrel} does not
include cases for variables. Likewise, for the base types such as
boolean and equality, the output set $[[S]]$ contains only terms that
evaluate to closed terms.

\scw{What is the motivation for this section? Why do we want to generalize
the logical relation?}
To generalize our logical relation to open
terms, we define the semantic typing judgment by closing the open
terms with a substitution whose codomain consists of terms that
respect the interpretation of the types from the context. The full
definitions of well-formed substitution ($[[rho |= G]]$), semantic
typing ($[[ G |= a : A]]$), and semantic context well-formedness
($[[|= G]]$) are presented in Figure~\ref{fig:semtyping}.
\scw{explain all parts of this figure. it's pretty intimidating.}
In the
definition of $[[|= G]]$, we use the notation $[[drop i G]]$ to denote
the typing context obtained by dropping the last $[[i]]$ elements of
$[[G]]$. When $[[G]]$ has less than $[[i]]$ elements, $[[drop i G]]$
returns the empty list.

@@

The predicates $[[wne a]]$ and $[[wn
a]]$ describe terms that can evaluate into $\beta$-neutral or
$\beta$-normal form through parallel reduction and are defined as
follows.\scw{What does $wne$ stand for?}
\[ [[wne a]] \iff \exists [[e]], [[a =>+ e]] \]
\[ [[wn a]] \iff \exists [[f]], [[a =>+ f]] \]

@@

The updated logical relation and its auxiliary definitions are shown
in Figure~\ref{fig:logrelopen}.

% The idea of a term blocked from $\beta$ reduction is
% captured by $[[e]]$, the set of neutral terms. We augment the
% interpretation of booleans and the empty types with terms that
% evaluate to neutral terms. Furthermore, types that are neutral terms
% can also be assigned a meaning since they may be inhabited by other
% neutral terms.

In the definition of the logical relation, we omit the rules for
the function and universe cases since they remain identical to the
original version in Figure~\ref{fig:logrel}.
The changes to \rref{I-Bool} and \rref{I-Void} follow the exact same pattern.
An open term of type $[[Bool]]$ does
not necessarily reduce to $[[true]]$ or $[[false]]$, but may reduce to
a variable, or more generally, a neutral term. Likewise, the
$[[Void]]$ type, while remains uninhabited under an empty context, may
be inherited by the set of neutral terms when there is a variable in
the context that allows us to inhabit $[[Void]]$.

>>>>

% The logical relation we define in Figure~\ref{fig:logrel} does not
% include cases for variables. Likewise, for the base types such as
% boolean and equality, the output set $[[S]]$ contains only terms that
% evaluate to closed terms.

% \scw{What is the motivation for this section? Why do we want to generalize
% the logical relation?}
In this section, we define the notion of semantic typing based on the
logical relation we have defined in Section~\ref{sec:logreldep} and
prove the fundamental lemma, which states that syntactic typing
implies semantic typing.

The necessity of semantic typing as an extra layer of definition on
top of the logic relation can be understood in simply typed
languages~\citep{skorstengaard2019introduction, harpertait,
  pierce2002types}. In our setting, attempting to show that $[[empty
|- a : A]]$ implies $[[InterpR i A S]]$ and $[[a in S]]$ will fail in
\rref{T-Abs}, where induction hypothesis is not helpful since the body
of the lambda term is typed under a non-empty context. Through
the definition of semantic typing, we can state a strengthened
property that is actually provable.

% To generalize our logical relation to open
% terms, we define the semantic typing judgment by closing the open
% terms with a substitution whose codomain consists of terms that
% respect the interpretation of the types from the context.
The full
definitions of well-formed substitution ($[[rho |= G]]$), semantic
typing ($[[ G |= a : A]]$), and semantic context well-formedness
($[[|= G]]$) are presented in Figure~\ref{fig:semtyping}.

The $[[rho |= G]]$ judgment defines the semantic well-formedness of
a substitution $[[rho]]$ with respect to a context
$[[G]]$. Intuitively, for variable $[[i]]$ in the context (thus the
restriction $[[i < | G |]]$), $[[rho i]]$ is a term that inhabits all
possible interpretations of the type $[[G i]]$ from the
context. However, since we are in a dependently typed system, there
may be variables in $[[G i]]$, too. The type $[[(G i < up Suc i > ) {
  rho }]]$ applies $[[rho]]$ to replace those variables. The shifting
operation $[[up Suc i]]$ can be viewed as an artifact of the de Bruijn
representation to lift the type $[[G i]]$ to the right scope before
substitution can be carried out.

In the
definition of $[[|= G]]$, we use the notation $[[drop i G]]$ to denote
the typing context obtained by dropping the last $[[i]]$ elements of
$[[G]]$. When $[[G]]$ has less than $[[i]]$ elements, $[[drop i G]]$
returns the empty list.

@@

The predicates $[[wne a]]$ and $[[wn
a]]$ describe terms that can evaluate into $\beta$-neutral or
$\beta$-normal form through parallel reduction and are defined as
follows.
\[ [[wne a]] \iff \exists [[e]], [[a =>+ e]] \]
\[ [[wn a]] \iff \exists [[f]], [[a =>+ f]] \]
We can read $[[wne a]]$ and $[[wn a]]$ as weakly normalizing to a neutral or a normal form respectively.

@@

The updated logical relation and its auxiliary definitions are shown
in Figure~\ref{fig:logrelopen}.
% The idea of a term blocked from $\beta$ reduction is
% captured by $[[e]]$, the set of neutral terms. We augment the
% interpretation of booleans and the empty types with terms that
% evaluate to neutral terms. Furthermore, types that are neutral terms
% can also be assigned a meaning since they may be inhabited by other
% neutral terms.
In the definition of the logical relation, we omit the rules for
the function and universe cases since they remain identical to the
original version in Figure~\ref{fig:logrel}.
The changes to \rref{I-Bool} and \rref{I-Void} follow the exact same pattern.
An open term of type $[[Bool]]$ does
not necessarily reduce to $[[true]]$ or $[[false]]$, but may reduce to
a variable, or more generally, a neutral term. Likewise, the
$[[Void]]$ type, while remains uninhabited under an empty context, may
be inherited by the set of neutral terms when there is a variable in
the context that allows us to inhabit $[[Void]]$.
